Ant cuticle image classification using texture analysis: a comparative study

### 2022-06-10 paper/main.tex
paper update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{txfonts}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{lipsum}
\usepackage{hyperref}
\hypersetup{colorlinks=true}
\usepackage{subcaption}
\usepackage{placeins}
\usepackage{tikz}
\usepackage{makecell}
\usetikzlibrary{automata,positioning,arrows}
\usepackage{caption}
\usepackage{changepage}

@@

\title{Ant Cuticle Texture Classification Algorithm for Ecological Anaylsis}

\author{%
    Noah Gardner\affil{1},
    John Paul Hellenbrand\affil{2},
    Anthony Phan\affil{1},
    Evan Zhu\affil{1},
    Zhiling Long\affil{3},
    and Chih-Cheng Hung\affil{1}\corrauth
}%

@@

\address{%
    \addr{\affilnum{1}}{Laboratory of Machine Vision and Security Research, %
        College of Computing and Software Engineering, Kennesaw State %
        University, Marietta GA, USA}%
    \addr{\affilnum{2}}{College of Science and Mathematics, Kennesaw State
        University, Kennesaw, GA, USA}%
    \addr{\affilnum{3}}{Department of Mathematics, Science, and Informatics,
        Mercer University, Macon, GA, USA}%
}
\corraddr{chung1@kennesaw.edu}

\editor{First-name Last-name}

\begin{abstract}
    There is a large variety of ant species, and most species are diverse in
    terms of size, shape, behaviors, and especially cuticle textures. However,
    the significance of ant cuticle texture is not widely researched. Ant
    cuticle texture presumably provides some type of function, and therefore is
    useful to research for ecological applications and bioinspired designs. This
    research employs modern machine learning methods such as texture analysis
    and deep machine learning to automatically group similar ant species based
    on morphological traits. We provide a comparative study of the performance
    of modern texture analysis methods on ant head images. We evaluate the
    results of the classification methods with modern visualization techniques.
\end{abstract}
\keywords{texture analysis; image processing; clustering; machine learning; ecology}
\maketitle

\section{Introduction}
Insects compose half of biodiversity and rank among the most dominant organisms
in terrestrial ecosystems \cite{sheikh_diverse_2017}. A key factor for the
ecological success of insects is their exoskeleton, also known as cuticle. The
cuticle protects insects from predation, provides structural support, prevents
desiccation, and serves as a canvas for advertising visual and chemical signals
\cite{gullan_insects_2009}. Research has heavily focused on the macrostructures
and internal chemical components that make the exoskeleton functional and more
recent work is being done to understand the functional aspects of external
cuticle micro sculpturing \cite{muthukrishnan_insect_2020,
    gunderson_insect_1989, watson_diversity_2017}.

Due to the extensive number of insect species, manual exploration of
insect-based information is difficult and often requires specialized expertise.
Therefore, automated entomology is gaining attraction by both biologists and
computer scientists and is expected to be a major contribution to the future of
insect-based research \cite{martineau_survey_2017}. One of the most commonly
used data types for insect analysis is image data. To develop an image-based
system for insect analysis, we can take advantage of existing work in general
image processing and texture analysis methods.

We examine ants (\textit{Formicidae}) as they display an extreme diversity of
cuticle micro sculpturing across all subfamilies. Sculpturing ranges from
parallel longitudinal ridges to deep oval impressions to erratic protuberances.
The sculpturing has arisen convergently and independently throughout ant's
evolutionary history, which suggests some inherent function. Cuticle sculpturing
on ants may help increase strength and rigidness, resist abrasion, increase
internal and external surface area, resist microbial growth, and rear beneficial
anti-biotic producing bacteria \cite{johnson_effect_2011,
    bruckner_relationship_2017, currie_coevolved_2006}. These specific functions may
be associated with certain sculpturing types and the purpose of classification
is to group similar textures based on proposed function.

In many texture analysis methods, the general goal is to automatically
categorize an object into a set of objects with similar texture-based features.
The approach of texture analysis to categorize similarly texture-based objects
corresponds well to the demands of ant identification. Texture analysis has
shown promising results in related fields, such as plant identification
\cite{boudra_plant_2018}. With modern texture analysis methods, the
categorization of ants can be automated and the results can be used to study the
influence of cuticle texture on ant ecology.

\section{Related Work}
\subsection{Sculpturing Identification}
Taxonomists have developed extensive terminology describing ant cuticle
sculpturing as it is often a useful diagnostic trait to distinguish between
closely related ant species \cite{blaimer_taxonomy_2019,fisher_ants_2007}. The
definitive text on ant cuticle terminology — \textit{The Glossary of Surface
    Sculpturing} — contains over 100 terms to describe the cuticle sculpturing
patterns of ants \cite{harris_glossary_1979}. There is often substantial overlap
among terms, and closely related cuticle patterns are likely to be functionally
similar. For example, the definition for “imbricate” is, “partly overlapping and
appearing like shingles on a roof or scales on a fish,” which is difficult to
distinguish from the definition of tesselate, “made up of squares like a chess
board, either in sculpturing or in color.”

Cuticle sculpturing in ants has been explored thoroughly from a taxonomic
perspective; however, the function of nano and microstructures on insect
exoskeletons is a developing topic in entomology. Watson et al. reviews the
literature of cuticle nano and microstructure function and proposes 21 possible
functions associated with these structures \cite{watson_diversity_2017}. Many of
these functions are related to structures not found in ants such as scales and
nanostructures. The review does include functions that may relate to ants such
as friction control, enhanced surface area, and increased hardness.  Watson et
al. also describe seven types of cuticle structures ranging from hairs and
scales to nano and micro structures \cite{watson_diversity_2017}. The cuticle
sculpturing of ants seems to fall within one type - complex microstructures.

The five broad functional groupings developed that describes the complex
microstructures found on ants were derived from reviewing the variation of
cuticle sculpturing across ants. These functional groupings were also reflected
in Harris as many terms could be grouped together based on similar definitions
and comparing the SEM photographs provided in the publication
\cite{harris_glossary_1979}.

\subsection{Insect Classification}

In this section, we provide an overview of some insect classification methods.
Proposed insect classification methods seek to classify insects at different
hierarchical levels, such as species, genus, family, and order. Additionally,
some methods may classify insects at a combination of different hierarchical
levels. Insect classification methods can be applied to a variety of fields. In
agriculture, insect classification methods can be used to identify the presence
of pest insects in crops, which can inform crop managers in their choice of
pesticides and help prevent crop loss \cite{liu_pestnet_2019,
    kasinathan_machine_2021}.

Feng et al. \cite{feng_automated_2013} apply an automated system to classify
moth images based on semantic related visual attributes, which are defined as a
pattern on the moth wings. Feng et al. \cite{feng_automated_2013} use a custom
texture descriptor based on the combination of Grey Level Co-occurence Matrtix
and \textit{Scale-Invariant Feature Transform} (SIFT) features
\cite{gotlieb_texture_1990, lowe_distinctive_2004}. The method proposed by Feng
et al. \cite{feng_automated_2013} is used to classify 50 different moth species
across 8 families and is based on standard texture analysis features
\cite{feng_automated_2013}. The results from Feng et al.
\cite{feng_automated_2013} suggest that traditional feature extraction
techniques for the semantic visual attributes of the moth wings are sufficient
for training a classifier to classify an image between 10 randomly selected moth
species.

Urteaga et al. \cite{urteaga_scorpions_2016} use machine learning methods in
order to classify images between two different scorpion species:
\textit{Centruroides limpidus} and \textit{Centruroides noxius}. After applying
background distinction based on dynamic color threshold, Urteaga et al.
\cite{urteaga_scorpions_2016} apply feature extraction to extract features from
the separated scorpion image such as aspect ratio, rectangularity, and
compactness. Urteaga et al. \cite{urteaga_scorpions_2016} apply three different
classification models to classify the image as one of the species: Artificial
Neural Network, Regression Tree, and Random Forest classifiers
\cite{urteaga_scorpions_2016}. The results from Urteaga et al.
\cite{urteaga_scorpions_2016} show that after background removal,
characteristics from the entire body of the scorpion can be used to create a
binary classifier that can classify the image as one of the two species.
Therefore, we may consider an experiment on the presence of the background from
the ant head image in order to improve classification accuracy. However, the
features extracted from \cite{urteaga_scorpions_2016} do not make use of any
deep learning methods and are not compatible with our texture-based dataset.

Lim et al. \cite{lim_performance_2017} apply a CNN-based algorithm for insect
classification. Lim et al. \cite{lim_performance_2017} classify a subset of
insect species and families based on the classes available in the ImageNet
dataset. ImageNet is a widely used dataset of images labeled by experts with
millions of images and thousands of categories \cite{deng_imagenet_2009}. In the
ImageNet dataset, there are some categories that specify the class of the insect
on a species level, \textit{e.g.} \textit{monarch butterfly} and \textit{ringlet
    butterfly} as well as some categories that specify the class of the insect on a
family level, \textit{e.g.} \textit{ant}, \textit{fly}, and \textit{bee}
\cite{imagenet_labels}. Lim et al. \cite{lim_performance_2017} use a modified
AlexNet architecture and experiment with different numbers of kernels and their
effect the performance of the model. Glick et al. \cite{glick_insect_2016}
employ a similar approach by classifying 277 insect classes from ImageNet using
a hierarchical contains. The results from Lim et al. \cite{lim_performance_2017}
and Glick et al. \cite{glick_insect_2016} suggest that a CNN is capable of
differentiating between different hierarchical classes of insects. In our
research, we are interested in the classification of ant images of a
hierarchical level between species and family.

@@

\subsection{Dataset}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{thesis_assets/images/rough_collage.png}
    \caption{Examples of rough cuticle texture ant images in the dataset after
        center cropping.}
    \label{fig:rough-cuticle-texture}
\end{figure}

In this section, we describe the creation of the custom dataset used in this
research. In our dataset, we classify ant head images from AntWeb
\cite{perrichot_antweb_2012} into two categories: \textit{rough} and
\textit{smooth}. Some randomly selected images from each category are shown in
Figures \ref{fig:rough-cuticle-texture} and \ref{fig:smooth-cuticle-texture}.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{thesis_assets/images/smooth_collage.png}
    \caption{Examples of smooth cuticle texture ant images in the dataset after
        center cropping.}
    \label{fig:smooth-cuticle-texture}
\end{figure}

\subsubsection{Sculpture Identification Protocol}

To begin, a master spreadsheet was created with the 2,499 different ant species
to be identified for the primary dataset. A team of three assistants were in
charge of the manual identification process. The team was trained to identify
cuticle sculpturing through a process which consisted of one 45-minute
introductory lesson explaining the project and texture categories and then given
a training set of photos to identify from the genus \textit{Polyrhachis}, whose
members display among the highest diversity of cuticle patterns all categories.
The sculpture identification protocol describes the two primary categories:
\textit{rough} and \textit{smooth}.

@@

\subsubsection{Data Collection}

To collect the images, the assistants followed the taxonomy information
available in the master spreadsheet to the appropriate AntWeb page. In many
cases, there are multiple ant head images of the same species, and occasionally
there are multiple image resolution available from a single image. To simplify
the data collection process, the assistants were instructed to download the
first ant head image of the species being identified in the highest resolution
possible. Each image was named with an identifier that corresponds with the row
number in the master spreadsheet. The same ant head images that were downloaded
in the data collection phase were the same ones used in the sculpture
identification protocol. Ant species which did not have any images of the head
were excluded from the dataset. Additionally, ant species which only had a head
image of a queen ant were excluded from the dataset.

Ant specimen images taken from AntWeb \cite{perrichot_antweb_2012} are created
by different photographers and therefore have different attributes, such as
environment, resolution, and lighting. In the ant head images, the ant head is
in the center of the image and the body is pointing away from the camera. The
focus of the ant head image is centered on the head, with the background and
image artifacts from the ant body typically blurred. In most ant head images,
there is a bar which indicates the scale of the image due to the variety in the
sizes of different ant species. In a few ant head images, there exists some text
denoting the specimen identifier and other info. In terms of texture, some ant
specimens are very old, so their head images have other abnormalities such as
cracks in the cuticle and the presence of dust.

\subsubsection{Data Preprocessing}

@@

\subsection{Models}

@@

\subsubsection{Evaluation}

We evaluate the performance of the models according to standard evaluation
methods. Since we are working with a binary classification problem, we use a
standard confusion matrix to evaluate the accuracy, precision, and F1 score. We
also apply Grad-CAM with manual inspection to visualize the activation weights
for incorrectly classified images to determine which features are interfering
with the classification. Finally, we apply t-SNE to visualize the separation
learned for the model to further analyze the classifications made by the model.

@@

To begin, we share the results on the VGG model architectures on our custom
dataset. The classification results are summarized in a confusion matrix for
each model. Then, the statistics for each VGG architecture are shared in Table
\ref{tab:vgg_results} and Figure \ref{fig:vgg_results}. The results are
collected on 8 iterations of training and averaged. The results are
rounded to 2 decimal places where appropriate.

\vspace{0.5in}

\begin{table}[h]
    \centering
    \caption{VGG11 confusion matrix on ant head image dataset.}
    \begin{tabular}{r|ll}
                         & True Rough & True Smooth \\
        \hline
        Predicted Rough  & 101.75     & 53          \\
        Predicted Smooth & 37.375     & 257.875     \\
    \end{tabular}
    \label{tab:vgg11_confusion_matrix}
\end{table}

\vspace{0.5in}

\begin{table}[h]
    \centering
    \caption{VGG13 confusion matrix on ant head image dataset.}
    \begin{tabular}{r|ll}
                         & True Rough & True Smooth \\
        \hline
        Predicted Rough  & 95.25      & 39.375      \\
        Predicted Smooth & 43.25      & 272.25      \\
    \end{tabular}
    \label{tab:vgg13_confusion_matrix}
\end{table}

\vspace{0.5in}

\begin{table}[h]
    \centering
    \caption{VGG16 confusion matrix on ant head image dataset.}
    \begin{tabular}{r|ll}
                         & True Rough & True Smooth \\
        \hline
        Predicted Rough  & 87.5       & 42.625      \\
        Predicted Smooth & 47.375     & 272.5       \\
    \end{tabular}
    \label{tab:vgg16_confusion_matrix}
\end{table}

\vspace{0.5in}

\begin{table}[h]
    \centering
    \caption{VGG19 confusion matrix on ant head image dataset.}
    \begin{tabular}{r|ll}
                         & True Rough & True Smooth \\
        \hline
        Predicted Rough  & 91.25      & 42.125      \\
        Predicted Smooth & 43.25      & 273.375     \\
    \end{tabular}
    \label{tab:vgg19_confusion_matrix}
\end{table}

\vspace{0.5in}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\textwidth]{thesis_assets/plots/vgg_average.png}
    \caption{Average accuracy for VGG architectures on ant head image dataset.}
    \label{fig:vgg_results}
\end{figure}

@@

Next, we share the results on the ResNet model architectures on our custom
dataset with random weight initialization. The classification results are
summarized in a confusion matrix for each model. Then, the statistics for each
ResNet architecture are shared in Table \ref{tab:resnet_results} and Figure
\ref{fig:resnet_results}. The results are collected on 8 iterations of training and
averaged. The results are rounded to 2 decimal places where appropriate.

\begin{table}[h]
    \centering
    \caption{ResNet18 confusion matrix on ant head image dataset.}
    \begin{tabular}{r|ll}
                         & True Rough & True Smooth \\
        \hline
        Predicted Rough  & 98.75      & 48.75       \\
        Predicted Smooth & 39.25      & 262         \\
    \end{tabular}
    \label{tab:resnet18_confusion_matrix}
\end{table}

\vspace{0.5in}

\begin{table}[h]
    \centering
    \caption{ResNet50 confusion matrix on ant head image dataset.}
    \begin{tabular}{r|ll}
                         & True Rough & True Smooth \\
        \hline
        Predicted Rough  & 81.375     & 43.625      \\
        Predicted Smooth & 55.875     & 269.125     \\
    \end{tabular}
    \label{tab:resnet50_confusion_matrix}
\end{table}

\begin{table}[h]
    \centering
    \caption{ResNet101 confusion matrix on ant head image dataset.}
    \begin{tabular}{r|ll}
                         & True Rough & True Smooth \\
        \hline
        Predicted Rough  & 85.75      & 53.875      \\
        Predicted Smooth & 50.125     & 260.25      \\
    \end{tabular}
    \label{tab:resnet101_confusion_matrix}
\end{table}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\textwidth]{thesis_assets/plots/resnet_average.png}
    \caption{Average accuracy for ResNet architectures on ant head image dataset.}
    \label{fig:resnet_results}
\end{figure}

@@

Next, we share the results on the fine-tuned ResNet model architectures on our
custom dataset with pretrained weights. The classification results are
summarized in a confusion matrix for each model. Then, the statistics for each
ResNet architecture are shared in Table \ref{tab:fresnet_results} and Figure
\ref{fig:fresnet_results}. The results are collected on 8 iterations of training
and averaged. The results are rounded to 2 decimal places where appropriate.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{txfonts}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{lipsum}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{colorlinks=true}
\usepackage{subcaption}
\usepackage{placeins}
\usepackage{tikz}
\usepackage{makecell}
\usetikzlibrary{automata,positioning,arrows}
\usepackage{caption}
\usepackage{changepage}

@@

\title{Ant cuticle image classification using texture analysis: a comparative study}

\author{%
    Noah Gardner\affil{1},
    John Paul Hellenbrand\affil{2},
    Anthony Phan\affil{1},
    Evan Zhu\affil{1},
    Zhiling Long\affil{3},
    Min Wang\affil{2},
    Clint Penick\affil{2},
    and Chih-Cheng Hung\affil{1}\corrauth
}%

@@

\address{%
    \addr{\affilnum{1}}{Laboratory of Machine Vision and Security Research, %
        College of Computing and Software Engineering, Kennesaw State %
        University, Marietta GA, USA}%
    \addr{\affilnum{2}}{College of Science and Mathematics, Kennesaw State
        University, Kennesaw, GA, USA}%
    \addr{\affilnum{3}}{Department of Mathematics, Science, and Informatics,
        Mercer University, Atlanta, GA, USA}%
}
\corraddr{chung1@kennesaw.edu}

\editor{Pasi Fr\"{a}nti}

\begin{abstract}
    There is a large variety of ant species, and most species are diverse in
    terms of size, shape, behaviors, and especially cuticle textures. However,
    the significance of ant cuticle texture is not widely researched. Ant
    cuticle texture presumably provides some type of function, and therefore is
    useful to research for ecological applications and bioinspired designs. This
    research employs image texture analysis and deep machine learning to
    automatically group similar ant species based on morphological traits. We
    provide a comparative study of the performance of texture analysis methods
    on ant images. We evaluate the results of the classification methods with
    modern visualization techniques.
    \colorbox{green}{TODO: what is the result? experiments? data?}
\end{abstract}
\keywords{texture analysis; image processing; classification; machine learning;
    ant cuticle images; ecology}
\maketitle

\section{Introduction}
\colorbox{green}{TODO: refer to ant head images in a consistent way}

\colorbox{green}{TODO: refer to traditional/classical in a consistent way}

\colorbox{green}{TODO: change ideal and non-ideal to features identified by
    the biologist and to invalid features (background, etc.)}

Insects compose half of biodiversity and rank among the most dominant organisms
in terrestrial ecosystems \cite{sheikh_diverse_2017}. A key factor for the
ecological success of insects is their exoskeleton, also known as cuticle. The
cuticle protects insects from predation, provides structural support, prevents
desiccation, and serves as a canvas for advertising visual and chemical signals
\cite{gullan_insects_2009}. Research has heavily focused on the macrostructures
and internal chemical components that make the exoskeleton functional and more
recent work is being done to understand the functional aspects of external
cuticle micro sculpturing \cite{muthukrishnan_insect_2020,
    gunderson_insect_1989, watson_diversity_2017}.

Due to the extensive number of insect species, manual exploration of
insect-based information is difficult and often requires specialized expertise.
Therefore, automated entomology is gaining attraction by both biologists and
computer scientists and is expected to be a major contribution to the future of
insect-based research \cite{martineau_survey_2017}. One of the most commonly
used data types for insect analysis is image data. To develop an image-based
system for insect analysis, we can take advantage of existing work in general
image analysis methods.

We examine ants (\textit{Formicidae}) as they display an extreme diversity of
cuticle micro sculpturing across all subfamilies. Sculpturing ranges from
parallel longitudinal ridges to deep oval impressions to erratic protuberances.
The sculpturing has arisen convergently and independently throughout ant's
evolutionary history, which suggests some inherent function. Cuticle sculpturing
on ants may help increase strength and rigidness, resist abrasion, increase
internal and external surface area, resist microbial growth, and rear beneficial
anti-biotic producing bacteria \cite{johnson_effect_2011,
    bruckner_relationship_2017, currie_coevolved_2006}. These specific functions may
be associated with certain sculpturing types. In order to analyze those
functions, it is necessary to segment the image and group similar textures for
further analysis.

In image texture analysis methods, the general goal is to automatically
categorize an object into a set of objects with similar texture-based features.
Texture analysis has shown promising results in related fields, such as plant
identification \cite{boudra_plant_2018} \colorbox{green}{TODO: more references
    and examples}. With modern texture analysis methods, the categorization of ants
can be automated and the results can be used to study the influence of cuticle
texture on ant ecology.

Image texture analysis heavily depends on the spatial relationships among gray
levels of pixels \cite{hung_image_2019}. Hence, feature extraction method is
usually applied to the spatial relationship. The gray-level co-occurrence matrix
(GLCM) and local binary patterns (LBP) are two of methods which are based on the
spatial relationships to extracting features for this purpose
\cite{haralick_textural_1973, goos_gray_2000}. Traditional image texture
analysis can be divided into four categories: statistical, structural,
model-based, and transform-based methods \colorbox{green}{[references]}. In the
past decade, the deep learning network becomes the mainstream in image
classification and segmentation \cite{liu_bow_2019}. Compared with the
traditional method in which a kernel must be designed for extracting features,
the deep learning network can automatically extract the features by itself.

In this study, we used some of both approaches for our work on the ant images.

\colorbox{green}{TODO: Specifically, what approaches used? Why do we use image
    texture for ant images? What are the contributions?}

\colorbox{green}{TODO: briefly discuss results from the research}

\section{Related Work}

\colorbox{green}{TODO: introduce the section}

\subsection{Sculpturing Identification}
Taxonomists have developed extensive terminology describing ant cuticle
sculpturing as it is often a useful diagnostic trait to distinguish between
closely related ant species \cite{blaimer_taxonomy_2019,fisher_ants_2007}. The
definitive text on ant cuticle terminology — \textit{The Glossary of Surface
    Sculpturing} — contains over 100 terms to describe the cuticle sculpturing
patterns of ants \cite{harris_glossary_1979}. There is often substantial overlap
among terms, and closely related cuticle patterns are likely to be functionally
similar. For example, the definition for “imbricate” is, “partly overlapping and
appearing like shingles on a roof or scales on a fish,” which is difficult to
distinguish from the definition of tesselate, “made up of squares like a chess
board, either in sculpturing or in color.”

Cuticle sculpturing in ants has been explored thoroughly from a taxonomic
perspective; however, the function of nano and microstructures on insect
exoskeletons is a developing topic in entomology. Watson et al. review the
literature of cuticle nano and microstructure function and propose 21 possible
functions associated with these structures \cite{watson_diversity_2017}. Many of
these functions are related to structures not found in ants such as scales and
nanostructures. The review does include functions that may relate to ants such
as friction control, enhanced surface area, and increased hardness.  Watson et
al. also describe seven types of cuticle structures ranging from hairs and
scales to nano and micro structures \cite{watson_diversity_2017}. The cuticle
sculpturing of ants seems to fall within one type - complex microstructures.

The five broad functional groupings developed that describe the complex
microstructures found on ants were derived from reviewing the variation of
cuticle sculpturing across ants. These functional groupings were also reflected
in Harris as many terms could be grouped together based on similar definitions
and comparing the scanning electron microscope photographs provided in the
publication \cite{harris_glossary_1979}.

\colorbox{green}{TODO: how do we connect this section to image texture analysis?
    (why do we research this and why image texture) }

\subsection{Texture Analysis}
\colorbox{green}{TODO: A brief survey on image texture and cite Liu's
    IJCV2019 survey paper}

\subsection{Insect Classification}
In this section, we provide an overview of some insect classification methods.
Proposed insect classification methods seek to classify insects at different
hierarchical levels, such as species, genus, family, and order. Additionally,
some methods may classify insects at a combination of different hierarchical
levels. Insect classification methods can be applied to a variety of fields. In
agriculture, insect classification methods can be used to identify the presence
of pest insects in crops, which can inform crop managers in their choice of
pesticides and help prevent crop loss \cite{liu_pestnet_2019,
    kasinathan_machine_2021}.

Feng et al. \cite{feng_automated_2013} apply an automated system to classify
moth images based on semantic related visual attributes, which are defined as a
pattern on the moth wings. Feng et al. \cite{feng_automated_2013} use a custom
texture descriptor based on the combination of GLCM and \textit{scale-invariant
    feature transform} (SIFT) features \cite{gotlieb_texture_1990,
    lowe_distinctive_2004}. The method proposed by Feng et al.
\cite{feng_automated_2013} is used to classify 50 different moth species across
8 families \cite{feng_automated_2013}. The results from Feng et al.
\cite{feng_automated_2013} suggest that traditional feature extraction
techniques for the semantic visual attributes of the moth wings are sufficient
for training a classifier to classify an image between 10 randomly selected moth
species.

Urteaga et al. \cite{urteaga_scorpions_2016} use machine learning methods in
order to classify images between two different scorpion species:
\textit{Centruroides limpidus} and \textit{Centruroides noxius}. After applying
background distinction based on dynamic color threshold, Urteaga et al.
\cite{urteaga_scorpions_2016} apply feature extraction to extract features from
the separated scorpion image such as aspect ratio, rectangularity, and
compactness. Urteaga et al. \cite{urteaga_scorpions_2016} apply three different
classification models to classify the image as one of the species: Artificial
Neural Network, Regression Tree, and Random Forest classifiers
\cite{urteaga_scorpions_2016}. The results from Urteaga et al.
\cite{urteaga_scorpions_2016} show that after background removal,
characteristics from the entire body of the scorpion can be used to create a
binary classifier that can classify the image as one of the two species.

Lim et al. \cite{lim_performance_2017} apply a CNN-based algorithm for insect
classification. Lim et al. \cite{lim_performance_2017} classify a subset of
insect species and families based on the classes available in the ImageNet
dataset. ImageNet is a widely used dataset of images labeled by experts with
millions of images and thousands of categories \cite{deng_imagenet_2009}. In the
ImageNet dataset, there are some categories that specify the class of the insect
on a species level, \textit{e.g.} \textit{monarch butterfly} and \textit{ringlet
    butterfly} as well as some categories that specify the class of the insect on a
family level, \textit{e.g.} \textit{ant}, \textit{fly}, and \textit{bee}
\cite{imagenet_labels}. Lim et al. \cite{lim_performance_2017} use a modified
AlexNet architecture and experiment with different numbers of kernels and their
effect the performance of the model. Glick et al. \cite{glick_insect_2016}
employ a similar approach by classifying 277 insect classes from ImageNet using
a hierarchical convolutional neural network. The results from Lim et al.
\cite{lim_performance_2017} and Glick et al. \cite{glick_insect_2016} suggest
that a CNN is capable of differentiating between different hierarchical classes
of insects.

@@

\colorbox{green}{TODO: do not jump into subsection directly without any context}

\subsection{Dataset preparation} % preparation generation

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{thesis_assets/images/rough_collage.png}
    \caption{Examples of rough cuticle texture ant images in the dataset after
        center cropping, from AntWeb \cite{perrichot_antweb_2012}.}
    \label{fig:rough-cuticle-texture}
\end{figure}

In this section, we describe the creation of the custom dataset used in this
research. In our dataset, we use ant head images from AntWeb
\cite{perrichot_antweb_2012} and define two categories for them based on the
appearance of the cuticle texture: \textit{rough} and \textit{smooth}. Some
randomly selected images from each category are shown in Figures
\ref{fig:rough-cuticle-texture} and \ref{fig:smooth-cuticle-texture}.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{thesis_assets/images/smooth_collage.png}
    \caption{Examples of smooth cuticle texture ant images in the dataset after
        center cropping, from AntWeb \cite{perrichot_antweb_2012}.}
    \label{fig:smooth-cuticle-texture}
\end{figure}

% \subsubsection{Sculpture identification protocol}

To begin, a master spreadsheet was created with the 2,499 different ant species
to be identified for the primary dataset. The team was trained to identify
cuticle sculpturing through a process which consisted of one 45-minute
introductory lesson explaining the project and texture categories. Then, the
team was given a training set of photos to identify from the genus
\textit{Polyrhachis}. The sculpture identification protocol describes the two
primary categories: \textit{rough} and \textit{smooth}.

@@

% \subsubsection{Data collection}

To collect the images, the assistants followed the taxonomy information
available in the master spreadsheet to the appropriate AntWeb page
\cite{perrichot_antweb_2012}. In many cases, there are multiple ant head images
of the same species, and occasionally there are multiple image resolution
available from a single image. To simplify the data collection process, the
assistants were instructed to download the first ant head image of the species
being identified in the highest resolution possible. Each image was named with
an identifier that corresponds with the row number in the master spreadsheet.
The same ant head images that were downloaded in the data collection phase were
the same ones used in the sculpture identification protocol. Ant species which
did not have any images of the head were excluded from the dataset.
Additionally, ant species which only had a head image of a queen ant were
excluded from the dataset.

Ant specimen images taken from AntWeb \cite{perrichot_antweb_2012} are created
by different photographers and therefore have different attributes, such as
environment, resolution, and lighting. In the ant head images, the ant head is
in the center of the image and the body is pointing away from the camera. The
focus of the ant head image is centered on the head, with the background and
image artifacts from the ant body typically blurred. In most ant head images,
there is a bar which indicates the scale of the image due to the variety in the
sizes of different ant species. In a few ant head images, there exists some text
denoting the specimen identifier and other information. In terms of texture,
some ant specimens are very old, so their head images have other abnormalities
such as cracks in the cuticle and the presence of dust.

% \subsubsection{Data preprocessing}

@@

\subsection{Classical texture analysis methods}

\colorbox{green}{TODO: kviews, gabor, etc. will go here}

\subsection{Deep learning models used for the experiments}

@@

\colorbox{green}{\parbox{\dimexpr\textwidth-2\fboxsep}{TODO: I suggest that
        at least one deep learning based texture classification algorithm be applied
        to the dataset for the comparative study. When this is the case, one more
        subsection should be included here to describe that/those algorithm(s).}}

\subsection{Evaluation}

We evaluate the performance of the models according to standard evaluation
methods. Since we are working with a binary classification problem, we use a
standard confusion matrix to evaluate the accuracy, precision, and F1 score.
\colorbox{green}{(TODO: AUC and ROC can also be used as additional evaluation
    methods.)} We also apply Grad-CAM with manual inspection to visualize the
activation weights for classified images to visualize which features lead to the
classification result. Finally, we apply t-SNE to visualize the separation
learned for the model to further analyze the classifications made by the model.

@@

To begin, we share the results on the VGG model architectures on our custom
dataset. The classification results are summarized in a confusion matrix for
each model. Then, the statistics for each VGG architecture are shared in Table
\ref{tab:vgg_results}. The results are collected on 8 runs of training and
averaged. The results are rounded to 2 decimal places where appropriate.

% \begin{table}[h]
%     \centering
%     \caption{VGG11 confusion matrix on ant head image dataset.}
%     \begin{tabular}{r|ll}
%                          & True Rough & True Smooth \\
%         \hline
%         Predicted Rough  & 101.75     & 53          \\
%         Predicted Smooth & 37.375     & 257.875     \\
%     \end{tabular}
%     \label{tab:vgg11_confusion_matrix}
% \end{table}

% \begin{table}[h]
%     \centering
%     \caption{VGG13 confusion matrix on ant head image dataset.}
%     \begin{tabular}{r|ll}
%                          & True Rough & True Smooth \\
%         \hline
%         Predicted Rough  & 95.25      & 39.375      \\
%         Predicted Smooth & 43.25      & 272.25      \\
%     \end{tabular}
%     \label{tab:vgg13_confusion_matrix}
% \end{table}

% \begin{table}[h]
%     \centering
%     \caption{VGG16 confusion matrix on ant head image dataset.}
%     \begin{tabular}{r|ll}
%                          & True Rough & True Smooth \\
%         \hline
%         Predicted Rough  & 87.5       & 42.625      \\
%         Predicted Smooth & 47.375     & 272.5       \\
%     \end{tabular}
%     \label{tab:vgg16_confusion_matrix}
% \end{table}

% \begin{table}[h]
%     \centering
%     \caption{VGG19 confusion matrix on ant head image dataset.}
%     \begin{tabular}{r|ll}
%                          & True Rough & True Smooth \\
%         \hline
%         Predicted Rough  & 91.25      & 42.125      \\
%         Predicted Smooth & 43.25      & 273.375     \\
%     \end{tabular}
%     \label{tab:vgg19_confusion_matrix}
% \end{table}

@@

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=0.6\textwidth]{thesis_assets/plots/vgg_average.png}
%     \caption{Average accuracy for VGG architectures on ant head image dataset.}
%     \label{fig:vgg_results}
% \end{figure}

@@

Next, we share the results on the ResNet model architectures on our custom
dataset with random weight initialization. The classification results are
summarized in a confusion matrix for each model. Then, the statistics for each
ResNet architecture are shared in Table \ref{tab:resnet_results}. The results
are collected on 8 runs of training and averaged. The results are rounded
to 2 decimal places where appropriate.

% \begin{table}[h]
%     \centering
%     \caption{ResNet18 confusion matrix on ant head image dataset.}
%     \begin{tabular}{r|ll}
%                          & True Rough & True Smooth \\
%         \hline
%         Predicted Rough  & 98.75      & 48.75       \\
%         Predicted Smooth & 39.25      & 262         \\
%     \end{tabular}
%     \label{tab:resnet18_confusion_matrix}
% \end{table}

% \begin{table}[h]
%     \centering
%     \caption{ResNet50 confusion matrix on ant head image dataset.}
%     \begin{tabular}{r|ll}
%                          & True Rough & True Smooth \\
%         \hline
%         Predicted Rough  & 81.375     & 43.625      \\
%         Predicted Smooth & 55.875     & 269.125     \\
%     \end{tabular}
%     \label{tab:resnet50_confusion_matrix}
% \end{table}

% \begin{table}[h]
%     \centering
%     \caption{ResNet101 confusion matrix on ant head image dataset.}
%     \begin{tabular}{r|ll}
%                          & True Rough & True Smooth \\
%         \hline
%         Predicted Rough  & 85.75      & 53.875      \\
%         Predicted Smooth & 50.125     & 260.25      \\
%     \end{tabular}
%     \label{tab:resnet101_confusion_matrix}
% \end{table}

@@

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=0.6\textwidth]{thesis_assets/plots/resnet_average.png}
%     \caption{Average accuracy for ResNet architectures on ant head image dataset.}
%     \label{fig:resnet_results}
% \end{figure}

@@

Next, we share the results on the fine-tuned ResNet model architectures on our
custom dataset with pretrained weights. The classification results are
summarized in a confusion matrix for each model. Then, the statistics for each
ResNet architecture are shared in Table \ref{tab:fresnet_results}. The results
are collected on 8 runs of training and averaged. The results are rounded
to 2 decimal places where appropriate.
